Combinatorial interpretations of holonomic recurrences

### 2021-08-29 article/main.tex
Apply auto-grammar fixes

Every recursive structure can be described using generating function, which then can be transformed into holonomic recurrences. It allows us to quickly compute number of objects of given structure with of given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations, like adding some vertices to the smaller trees. Although, there is no known way of interpreting it in the language of the combinatorial structures.

Initial inspiration for the thesis came from publication \textit{A natural counting of lambda terms}\cite{zigzag}, which shows bijection between seemingly different tree structures, which are lambda trees, black-white trees nad zigzag free trees. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structure itself.

In the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose approach that helped us to find intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation.

@@

In the section\ \ref{sec:approaches} we show approaches to the problem that helped people achieving their results. All of them were tried in order to find the answer to the original problem of interpreting holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately none of them helped us enough to get the final result. They are proven to be helpful when we were stuck though as most of them are able to divide problem into smaller parts and detect corner cases.

Section\ \ref{sec:brutal_iterative_approach} presents our final approach to the original problem, ie.\ computer-aided iterative approach that allowed us to find the intermediate solution. We belive that it is able to provide us with the full solution, but we were not able to achieve that due to the limited time that was available to us.

At last, in section\ \ref{sec:further_work}, we present our view on what can be done with the problem in the future. We belive that one day the problem will be fully solved and more generic approach to interpreting holonomic recurrences will emerge.

@@

In order to make reading of this article easier, let's fix some notation and make some assumptions. Any deviation from this will be clearly stated. If you have trouble guessing what some letter or word means, this should be your primary place to look for the answer.

@@

Uncle of node \(v\) is defined as sibling of the parent of \(v\).

@@

\(\no\) is implicitly defined as an identifier of non-existing node (eg.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

@@

Virtual node is artificially defined node that does not affect the structure definition (eg.\ we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting number of children of the node, so the actual leaves are still called leaves).

@@

For binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is a right (left) child of its parent.

For binary tree, we say that node \(v\) has a right (left) parent \(u\) \(\textiff\) \(v\) is a left (right) child of \(u\).

\(\pointed{v}\) is implicitly defined as a pointed node \(v\). We can treat it as node with additional label \(\pointed{}\).

@@

\(\size{X}\) is implicitly defined as a size of the object \(X\).

eg.\ stands for Latin exampli gratia, which means \textit{for example}.

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define full binary tree recursively as a node having no children (ie.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

Note that in order to compute \(k\)-th element of the sequence, \(a_k\), we just need to find constant corresponding to indeterminate \(z^k\).

\begin{example}
    \label{ex:bin_gf}
    Having our recursive definition of the object, we can easily construct a generating function for it. Let's create generating function that will count number of full binary trees consisting of exactly \(k\) internal nodes.

@@

In order to get some intuition on how it works underneath, let's take a look on how we can find number of the full binary trees of size \(2\). We can expand the recurrence a bit:

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without the need of digging into details. The tool we will be using extensively in this work is Maple package called GFUN\cite{gfun}. If you want to learn more about algorithms used in the package, we strongly encourage reading the referenced article.

@@

Note that the sum can only take finite number of subsequent elements of the sequence \(A\) into the relation.

@@

    For our purposes we will often ignore the initial conditions, as they can be easily described by exact number of the objects. The most interesting part for us is the relation between the finite number consecutive elements.

@@

Combinatorial interpretation is just fancy name for something you probably know and use. Eg.\ if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\) you interpret it as number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, eg.\ we may want to find number of trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent.

Unfortunately, as for the most of auto-generated code, they almost never come with intuitive interpretation. They are just some recurrences, but there is no commonly known method to find, how they correspond to the structures they describe.

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as:
    \[\enc{T} = \enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\]
    ie.\ catenation of identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

@@

We can also define prefix and suffix of \(T\).

@@

In the article we define some operations using its notation. As pre-order notation is bijective, you can obtain the structure the notation describe by reversing the process of encoding.

@@

It is easy to see that tree of size \(n\) (ie. with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

We are going to create bijection between left and right side of the recurrence. We will be using term \textit{bijective language} to describe talking about the recurrence using the objects it describes and their transformations.

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \(B(n)\) is simply set of all full binary trees of size \(n\),
    \item \((n + 2) B(n + 1)\) is set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) B(n)\) is set of all full binary trees of size \(n\) with one of the nodes pointed (remember that the node can be internal node or leaf),
    \item \(2\) at the beginning of \(2 (2n + 1) B(n)\) means that we will be using \(2\) copies of the set defined in previous bullet.
\end{itemize}

At the moment left side of the recurrence describe set of trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformation to make the sets equal.

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), ie. trees of size \(n\) with single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\)

The only thing that's left for us to verify is if we generate all trees of size \(n + 1\) with pointed leaf. As from the recursion, we know that \(2 (2n + 1) B(n)\) is total number of trees with single leaf pointed, it is enough to check if we do not generate any duplicate using \(R_l\) and \(R_r\). It can happen in 2 ways:
\begin{enumerate}
    \item operation itself generates duplicates or
    \item 2 different operations generate the same tree.
\end{enumerate}

First case can also be split into 2 subcases. Without loss of the generality we can analyze just one of the operations, let it be \(R_l\). The second case is symmetric.
\begin{itemize}
    \item for some nodes \(u \neq v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item for some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

In the second case it is enough to notice that \(R_l\) generates trees with pointed node being a left child, and \(R_r\) generates trees with pointed node being right child. That immediately provides us with conclusion that we are generating no duplicates.

@@

Let's think, what would have changed if we introduce unary nodes to the binary case\cite{motzkin}. We will be counting edges. Note that tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary tree. We don't change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    Left-leaning binary tree\todo{find official name; skewed binary tree is something different} is a binary tree with following constraint added: if a node has right child, it also have a left child.
\end{definition}

The bijection between unary-binary and left-leaning binary trees is very simple. We just represent unary nodes as nodes with just left child of left-leaning binary tree. The other cases translate immediately to corresponding ones. Please note that this transformation keeps the size of the tree (ie.~number of edges).

@@

\begin{definition}
    Virtual leaf is a virtual node added to the node in place of non-existing child.
\end{definition}

As we have added \(n + 2\) virtual nodes, we have added as much virtual edges.

@@

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave can not have sibling being a actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node;
        \end{enumerate}
    \item I - Internal node.
\end{enumerate}

@@

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point any node, as in a tree having \(n - 1\) edges, we have \(n\) actual and \(n + 1\) virtual ones.

@@

Simply speaking, node can have zero or strictly more than one children.

@@

\begin{definition}
    Well-weighted binary plane tree is a weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if internal node has weight 2, then its right child is not a leaf.
    \end{itemize}
\end{definition}

@@

This may seem counterintuitive, as by default we want to describe how to generate larger case from the smaller ones, but it will make the interpretation simpler.

@@

To achieve trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3.\ and replace whole subtree with subtree of \(t''\) with its root pointed. In this way, we will loose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\):

@@

As you could see in the examples above, most of the solutions are based on the simple analysis of number of laves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorially some holonomic recurrence.

@@

It may seem counter-intuitive at first, that this changes anything crucial, but with a change of the size definition, the holonomic recurrence may complicate itself drastically making the problem more difficult than it should be.

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, eg.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common root and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates tree of size being sum of sizes of the components increased by 1. If you had defined the size as number of leaves or null-nodes, the size would be just plain sum of sizes of the smaller trees.
\end{itemize}

@@

As the most common trees are binary ones, it may be worth an effort to find the some structure based on binary trees that can represent original structure (see section~\ref{sub:unary_binary_trees}).

\subsection{Weights variablization}%
\label{sub:weights_variablization}

@@

Unfortunately, decreasing size of the classes increases number of the sizes we need to analyze to get full result.

@@

Analyzing such small cases is much easier, but on the other side, there are many more classes that have to be considered to get full interpretation.

@@

The non-linear one has very simple interpretation. We can take 2 smaller trees and join them using newly created common root.

@@

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such simple case as binary trees is almost trivial, especially if one has seen the solution once.

@@

If all elaborate methods has failed and the problem is still too big to interpret it by hand, a computer comes in handy.

@@

The definition above is quite popular, but introduces several problems. One of them is a problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable bound (or is it even free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

@@

As you can see, we managed to eliminate the arbitrary variables. We can also easily check if given variable is bound by checking its depth.

@@

\begin{definition}
    \(s\) is a node with only right child.
\end{definition}

\begin{definition}
    \(l\) is a node with only left child.
\end{definition}

@@

You can note that not every binary tree represents valid term.

\begin{example}
    \(\n\; \no\; \n\; \n\; \no\; \no\; \no\) does not represent valid lambda term.
\end{example}

@@

\(s\) represents successor which can be applied only to \(n \in \N\), but is applied to lambda-abstracted term.

@@

As we have our structure defined, let's create generating functions that will keep track of more than just number of nodes, but also counts of each kind of nodes in the tree. The generating function will be parameterized then using following arguments:
\begin{itemize}
    \item \(z\) counting total number of virtual leaves
    \item \(o\) counting number of bound variable nodes
    \item \(l\) counting number of lambda abstraction nodes
    \item \(a\) counting number of application nodes
    \item \(s\) counting number of successors of bound variable nodes
\end{itemize}

@@

Using a little bit of Maple magic\cite{gfun}, we can easily achieve following holonomic recurrence.

@@

Appearing in the recurrence node kinds mean that we extend trees of given size using nodes of that kinds.

@@

You may notice that some potentially trivial terms are more complex that it seems. If we didn't count different types of nodes, we would achieve following holonomic recurrence:

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, ie.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, ie.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, ie.~is right child of its parent;
    \item node has its parent on the right, ie.~is left child of its parent.
\end{itemize}

@@

\begin{definition}
    Class \(T_i\) is defined as set of transformations of trees of size \((n - i)\) into trees of size \(n\).
\end{definition}

\begin{definition}
    Section \(T_{i,j}\) is defined as subset of \(T_i\) adding specific set of nodes. We assign \(j\) according to the lexicographical order of lexicographically sorted node types.
\end{definition}

\begin{definition}
    Part \(T_{i,j,k}\) is defined as single transformation in section \(T_{i,j}\). \(k\) is assigned according to the order of discovering the transformation.
\end{definition}

@@

\begin{definition}
    In context of \(T_i\), \(k\) is defined as \((n - i)\).
\end{definition}

\begin{definition}
    In context of \(T_i\), \(C_k\) is defined as \(L(n - i)\), ie.\ number of trees transformations from \(T_i\) are working on.
\end{definition}

\begin{definition}
    Section equation \(E_{i, j}\) is an equation defined in context of \(T_i\) that evaluates to a total number of trees that should be generated by transformations from \(T_{i, j}\).
\end{definition}

\begin{definition}
    Section expression \(S_{i, j}\) is pair of section equation and set of kinds of nodes the transformation introduces.
\end{definition}

@@

Using context-based variables makes the equation simpler to interpret, as we can immediately see relation between \(k\) and multiplier.

@@

As we want to make our work kind of organized, we are going to divide the process into iterations. Each iteration should be a small step into direction of solving the original problem.

We are going to generate two kinds of trees. Positive and negative ones. When computing total number of trees, we count positive tree is counted as \(1\), negative as \(-1\). Our goal is to generate \(E_{i, j}\) trees for each section \(T_{i, j}\).

@@

Please note that it is only an example of missing construction. It is just the simplest one we could think about, and in that case, simple solution is most often the correct one.

>>>>

Every recursive structure can be described using a generating function, which can be then transformed into holonomic recurrence. It allows us to quickly count objects of given structure with of given size. At first glance, the recurrence seems to be closely related to the structure itself. It just seems like it should describe simple transformations, like adding some vertices to the smaller trees. Although, there is no known method of interpreting it in the language of the combinatorial structures.

Initial inspiration for the thesis came from publication \textit{A natural counting of lambda terms}\cite{zigzag}, which shows bijection between seemingly different tree structures, which are lambda trees, black-white trees and zigzag free trees. They share the same holonomic recurrence, but there is no known immediate connection between the holonomic recurrence and the structure itself.

In the thesis we try to accumulate known results and techniques that may help solving the problem of finding combinatorial interpretations. We also propose an approach that helped us to find an intermediate solution for the problem of finding combinatorial interpretation for holonomic recurrence of lambda trees in the unary de Bruijn notation.

@@

In the section\ \ref{sec:approaches} we show approaches to the problem that helped people achieve their results. All of them were tried in order to find the answer to the original problem of interpreting holonomic recurrence of trees of lambda terms in the unary de Bruijn notation. Unfortunately none of them helped us enough to get the final result. They are proven to be helpful when we are stuck though as most of them are able to divide problems into smaller parts and detect corner cases.

Section\ \ref{sec:brutal_iterative_approach} presents our final approach to the original problem, ie.\ computer-aided iterative approach that allowed us to find the intermediate solution. We believe that it is able to provide us with the full solution, but we were not able to achieve that due to the limited time that was available to us.

Finally, in section\ \ref{sec:further_work}, we present our view on what can be done with the problem in the future. We believe that one day the problem will be fully solved and a more generic approach to interpreting holonomic recurrences will emerge.

@@

In order to make reading this article easier, let's fix some notation and make some assumptions. Any deviation from this will be clearly stated. If you have trouble guessing what some letter or word means, this should be your primary place to look for the answer.

@@

Uncle of node \(v\) is defined as a sibling of the parent of \(v\).

@@

\(\no\) is implicitly defined as an identifier of a non-existing node (eg.\ if \(v\) is a leaf, \(\tleft(v) = \n\)).

@@

Virtual node is an artificially defined node that does not affect the structure definition (eg.\ we can define virtual leafs that are children of actual leaves of the binary tree. Such virtual leaves are not included when counting the number of children of the node, so the actual leaves are still called leaves).

@@

For a binary tree, we say that node \(v\) is right (left) \(\textiff\) \(v\) is a right (left) child of its parent.

For a binary tree, we say that node \(v\) has a right (left) parent \(u\) \(\textiff\) \(v\) is a left (right) child of \(u\).

\(\pointed{v}\) is implicitly defined as a pointed node \(v\). We can treat it as a node with an additional label \(\pointed{}\).

@@

\(\size{X}\) is implicitly defined as the size of the object \(X\).

eg.\ stands for Latin exempli gratia, which means \textit{for example}.

@@

We will be analyzing structures that can be expressed using recursive formulas. Eg.\ we can define a full binary tree recursively as a node having no children (ie.\ being a leaf) or having exactly two subtrees being full binary trees. To express the alternative in the construction rule, we use \(+\). The reason will become obvious soon.

@@

Note that in order to compute \(k\)-th element of the sequence, \(a_k\), we just need to find a constant corresponding to indeterminate \(z^k\).

\begin{example}
    \label{ex:bin_gf}
    Having our recursive definition of the object, we can easily construct a generating function for it. Let's create a generating function that will count the number of full binary trees consisting of exactly \(k\) internal nodes.

@@

In order to get some intuition on how it works underneath, let's take a look at how we can find the number of full binary trees of size \(2\). We can expand the recurrence a bit:

@@

There exist multiple well-known methods and tools that allow us to work with generating functions without the need of digging into details. The tool we will be using extensively in this work is the Maple package called GFUN\cite{gfun}. If you want to learn more about algorithms used in the package, we strongly encourage reading the referenced article.

@@

Note that the sum can only take a finite number of subsequent elements of the sequence \(A\) into the relation.

@@

    For our purposes we will often ignore the initial conditions, as they can be easily described by the exact number of the objects. The most interesting part for us is the relation between the finite number consecutive elements.

@@

Combinatorial interpretation is just a fancy name for something you probably know and use. Eg.\ if you know the binomial coefficient \(\binom{n}{k}\), I bet it was introduced to you as something like: "let's assume you have \(n\) distinguishable balls. If you want to choose \(k\) from them, then you can do it in \(\binom{n}{k}\) ways" and now, whenever you see the term \(\binom{n}{k}\) you interpret it as number of ways to choose \(k\) objects from the set of size \(n\). This is the combinatorial interpretation of the binomial coefficient.

Holonomic recurrences are very useful if we want to count objects of given size. Of course, size can be defined in more complex ways, eg.\ we may want to find the number of trees with \(n\) internal nodes having exactly \(m\) nodes being left child of its parent.

Unfortunately, as for most auto-generated code, they almost never come with intuitive interpretation. They are just some recurrences, but there is no commonly known method to find how they correspond to the structures they describe.

@@

\begin{definition}
    Pre-order notation of \(n\)-ary tree \(T\), \(\enc{T}\), is defined recursively as:
    \[\enc{T} = \enc{\troot(T)} \prod_{i=1}^{n} \enc{i\text{-th subtree of } \troot(T)}\]
    ie.\ catenation of the identifier of the root of the tree with encoded its subtrees from left to right.
\end{definition}

@@

We can also define prefixes and suffixes of \(T\).

@@

In the article we define some operations using its notation. As pre-order notation is bijective, you can obtain the structure the notation describes by reversing the process of encoding.

@@

It is easy to see that a tree of size \(n\) (ie. with \(n\) internal nodes) has \(n + 1\) leaves and therefore \(2n + 1\) nodes in total. Analogically the tree of size \(n + 1\) has \(n + 2\) leaves and \(2n + 3\) nodes in total. If you see some resemblance to our holonomic recurrence, you are perfectly right.

We are going to create a bijection between the left and right side of the recurrence. We will be using the term \textit{bijective language} to talk about the recurrence using the objects it describes and their transformations.

Let's translate the recurrence to the bijective language then.
\begin{itemize}
    \item \(B(n)\) is simply set of all full binary trees of size \(n\),
    \item \((n + 2) B(n + 1)\) is set of all full binary trees of size \(n + 1\) with one of the leaves pointed,
    \item \((2n + 1) B(n)\) is set of all full binary trees of size \(n\) with one of the nodes pointed (remember that the node can be internal node or leaf),
    \item \(2\) at the beginning of \(2 (2n + 1) B(n)\) means that we will be using \(2\) copies of the set defined in the previous bullet.
\end{itemize}

At the moment the left side of the recurrence describes a set of trees of size \(n + 1\), the right, on the other hand, of size \(n\). Therefore, we need to apply some transformation to make the sets equal.

@@

We can now take the bijective interpretation of \((2n + 1) B(n)\), ie. trees of size \(n\) with a single node pointed. For each of such trees, \(T\) we create new one by applying \(R_l\) and \(R_r\) to the pair \(T', v\), where \(T'\) is \(T\) with removed pointers and \(v\) is a node that is pointed in \(T\). Therefore we generate \(2(2n + 1) B(n)\) trees of size \(n + 1\)

The only thing that's left for us to verify is if we generate all trees of size \(n + 1\) with pointed leaves. As from the recursion, we know that \(2 (2n + 1) B(n)\) is total number of trees with single leaf pointed, it is enough to check if we do not generate any duplicate using \(R_l\) and \(R_r\). It can happen in 2 ways:
\begin{enumerate}
    \item operation itself generates duplicates or
    \item 2 different operations generate the same tree.
\end{enumerate}

First case can also be split into 2 subcases. Without loss of generality we can analyze just one of the operations, let it be \(R_l\). The second case is symmetric.
\begin{itemize}
    \item for some nodes \(u \neq v\), \(R_l(T, u) = R_l(T, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) and \(\enc{v}\) in \(\enc{T}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(T, v)}\). Due to the definition of \(R_l\), we get that \(\pointed{i}_u = i_u + 1\) and \(\pointed{i}_v = i_v + 1\), which implies that \(u = v\) what we assumed false.
    \item for some trees \(T \neq S\), \(R_l(T, u) = R_l(S, v)\). Let \(i_u, i_v\) be the indices of respectively \(\enc{u}\) in \(\enc{T}\) and \(\enc{v}\) in \(\enc{S}\). Let \(\pointed{i}_u, \pointed{i}_v\) be the indices of pointed node respectively in \(\enc{R_l(T, u)}\) and \(\enc{R_l(S, v)}\). If \(\pointed{i}_u \neq \pointed{i}_v\), this case is trivially false, so we can assume that \(\pointed{i}_u = \pointed{i}_v\). Therefore, again from the definition of \(R_l\), we have that \(i_u = i_v\). That implies that \(\enc{\tPre(T, u)} \enc{\tSub(u)} \enc{\tSuf(T, u)} = \enc{\tPre(S, v)} \enc{\tSub(v)} \enc{\tSuf(S, v)} \), which then sums up to \(\enc{T} = \enc{S}\), which we assumed false.
\end{itemize}

In the second case it is enough to notice that \(R_l\) generates trees with pointed nodes being a left child, and \(R_r\) generates trees with pointed nodes being right child. That immediately provides us with the conclusion that we are generating no duplicates.

@@

Let's think, what would have changed if we introduce unary nodes to the binary case\cite{motzkin}. We will be counting edges. Note that a tree having \(n\) edges has exactly \(n + 1\) nodes.

@@

The definition above is precise, but introduces a new kind of node that we would have to take care of. We can avoid it by using a trick to make the tree recall binary trees. We don't change the structure, just the notation.

\begin{definition}
    \label{def:unary_binary_2}
    Left-leaning binary tree\todo{find official name; skewed binary tree is something different} is a binary tree with the following constraint added: if a node has a right child, it also has a left child.
\end{definition}

The bijection between unary-binary and left-leaning binary trees is very simple. We just represent unary nodes as nodes with just the left child of the left-leaning binary tree. The other cases translate immediately to corresponding ones. Please note that this transformation keeps the size of the tree (ie.~number of edges).

@@

\begin{definition}
    Virtual leaf is a virtual node added to the node in place of a non-existing child.
\end{definition}

As we have added \(n + 2\) virtual nodes, we have added as many virtual edges.

@@

\begin{enumerate}
    \item Vro - virtual leaf being right virtual-oneling;
    \item Vrt - virtual leaf being right virtual-twinling. We will divide this case into:
        \begin{enumerate}
            \item VrtUrv - Vri which uncle is a right virtual leaf;
            \item VrtUra - Vri which uncle is a right actual node;
            \item VrtUla - Vri which uncle is a left actual node.
        \end{enumerate}
    \item Vlt - virtual leaf left virtual-twinling. Note that from definition \ref{def:unary_binary_2}, left virtual leave can not have a sibling being an actual node. We can divide also this case into:
        \begin{enumerate}
            \item VltUrv - Vlt which uncle is a right virtual leaf;
            \item VltUra - Vlt which uncle is a right actual node;
            \item VltUla - Vlt which uncle is a left actual node;
        \end{enumerate}
    \item I - Internal node.
\end{enumerate}

@@

Let's start from \(U(n-1)\). Its multiplier is \(2n + 1\). It suggests that we can point to any node, as in a tree having \(n - 1\) edges, we have \(n\) actual and \(n + 1\) virtual ones.

@@

Simply speaking, nodes can have zero or strictly more than one child.

@@

\begin{definition}
    Well-weighted binary plane tree is a weighted tree such that:
    \begin{itemize}
        \item leaves are unweighted;
        \item internal nodes can have weight either \(1\) or \(2\);
        \item if the internal node has weight 2, then its right child is not a leaf.
    \end{itemize}
\end{definition}

@@

This may seem counterintuitive, as by default we want to describe how to generate larger cases from the smaller ones, but it will make the interpretation simpler.

@@

To achieve trees from \((n - 2) T(n - 1)\), we can simply take trees from case 3.\ and replace whole subtree with subtree of \(t''\) with its root pointed. In this way, we will lose \(2\) leaves and obtain each tree from \((n - 2) T(n - 1)\):

@@

As you could see in the examples above, most of the solutions are based on the simple analysis of number of leaves, internal nodes and other similar properties. In this section, we've gathered some methods that may be helpful when trying to analyze the problem of interpreting combinatorially some holonomic recurrence.

@@

It may seem counterintuitive at first, that this changes anything crucial, but with a change of the size definition, the holonomic recurrence may complicate itself drastically making the problem more difficult than it should be.

There are 2 most common symptoms that we may want to choose different method of counting the size:
\begin{itemize}
    \item we get empty classes of the objects, eg.\ if we analyze problem of full binary trees and define size as number of all existing nodes, we can not generate tree of even size, therefore half of our classes are empty;
    \item we want to join trees using common roots and we get many \(\pm1\) in our notes. Note that if you count internal nodes, such catenation creates a tree of size being the sum of sizes of the components increased by 1. If you had defined the size as the number of leaves or null-nodes, the size would be just a simple sum of the sizes of the smaller trees.
\end{itemize}

@@

As the most common trees are binary ones, it may be worth an effort to find some structure based on binary trees that can represent the original structure (see section~\ref{sub:unary_binary_trees}).

\subsection{Weights variabilization}%
\label{sub:weights_variablization}

@@

Unfortunately, decreasing the size of the classes increases the number of the sizes we need to analyze to get a full result.

@@

Analyzing such small cases is much easier, but on the other hand, there are many more classes that have to be considered to get full interpretation.

@@

The non-linear one has a very simple interpretation. We can take 2 smaller trees and join them using newly created common root.

@@

Using our inductive hypothesis, ie.\ holonomic recurrence and moving problematic cases out from the sums (in that case \(B(1)\) needs to be analyzed separately, as we did not define binary trees with no leaves), we can get the right side to use \(B(n - 1)\) and prove that the holonomic recurrence using non-linear one. The only part that is left is to translate this proof to the bijective language, which in such a simple case as binary trees is almost trivial, especially if one has seen the solution once.

@@

If all elaborate methods have failed and the problem is still too big to interpret it by hand, a computer comes in handy.

@@

The definition above is quite popular, but introduces several problems. One of them is a problem of potential collision of lambda-abstracted variables and difficulty of finding to which lambda abstraction is the variable bound (or is it even a free variable, ie.\ not bound). To address that, we can use de Bruijn indices.

@@

As you can see, we managed to eliminate the arbitrary variables. We can also easily check if a given variable is bound by checking its depth.

@@

\begin{definition}
    \(s\) is a node with only the right child.
\end{definition}

\begin{definition}
    \(l\) is a node with only the left child.
\end{definition}

@@

You can note that not every binary tree represents a valid term.

\begin{example}
    \(\n\; \no\; \n\; \n\; \no\; \no\; \no\) does not represent a valid lambda term.
\end{example}

@@

\(s\) represents the successor which can be applied only to \(n \in \N\), but is applied to the lambda-abstracted term.

@@

As we have our structure defined, let's create generating functions that will keep track of more than just the number of nodes, but also counts of each kind of node in the tree. The generating function will be parameterized then using following arguments:
\begin{itemize}
    \item \(z\) counting total number of virtual leaves
    \item \(o\) counting number of bound variable nodes
    \item \(l\) counting number of lambda abstraction nodes
    \item \(a\) counting number of application nodes
    \item \(s\) counting number of successors of bound variable nodes
\end{itemize}

@@

Using a little bit of Maple magic\cite{gfun}, we can easily achieve the following holonomic recurrence.

@@

Appearing in the recurrence node kinds means that we extend trees of given size using nodes of that kind.

@@

You may notice that some potentially trivial terms are more complex than they seem. If we didn't count different types of nodes, we would achieve following holonomic recurrence:

@@

We also created a framework\cite{transformator}, called transformator, that may help visualize our work. It is based on the idea that every transformation described using bijective language can be translated to a function transforming node and finite number of its surrounding nodes. This function can not discriminate between nodes on the grounds of their kind. The only discrimination criteria that may occur may be connected to the following cases:
\begin{itemize}
    \item node has no children, ie.~is a virtual leaf (which corresponds to the size of the tree, \(n\));
    \item node has has children, ie.~is not a virtual leaf (which corresponds to \(n - 1\));
    \item node is a root (which corresponds to \(1\));
    \item node has its parent on the left, ie.~is right child of its parent;
    \item node has its parent on the right, ie.~is the left child of its parent.
\end{itemize}

@@

\begin{definition}
    Class \(T_i\) is defined as a set of transformations of trees of size \((n - i)\) into trees of size \(n\).
\end{definition}

\begin{definition}
    Section \(T_{i,j}\) is defined as a subset of \(T_i\) adding a specific set of nodes. Webassign \(j\) according to the lexicographical order of lexicographically sorted node types.
\end{definition}

\begin{definition}
    Part \(T_{i,j,k}\) is defined as a single transformation in section \(T_{i,j}\). \(k\) is assigned according to the order of discovering the transformation.
\end{definition}

@@

\begin{definition}
    In the context of \(T_i\), \(k\) is defined as \((n - i)\).
\end{definition}

\begin{definition}
    In the context of \(T_i\), \(C_k\) is defined as \(L(n - i)\), ie.\ number of tree transformations from \(T_i\) are working on.
\end{definition}

\begin{definition}
    Section equation \(E_{i, j}\) is an equation defined in the context of \(T_i\) that evaluates to a total number of trees that should be generated by transformations from \(T_{i, j}\).
\end{definition}

\begin{definition}
    Section expression \(S_{i, j}\) is a pair of section equations and a set of kinds of nodes the transformation introduces.
\end{definition}

@@

Using context-based variables makes the equation simpler to interpret, as we can immediately see the relation between \(k\) and multiplier.

@@

As we want to make our work kind of organized, we are going to divide the process into iterations. Each iteration should be a small step into the direction of solving the original problem.

We are going to generate two kinds of trees. Positive and negative ones. When computing the total number of trees, we count positive trees as \(1\), negative as \(-1\). Our goal is to generate \(E_{i, j}\) trees for each section \(T_{i, j}\).

@@

Please note that it is only an example of missing construction. It is just the simplest one we could think about, and in that case, a simple solution is most often the correct one.
